High-Performance Language Primitives for Distributed Architectures using Asynchronous Events

### 2012-07-23 doc/asplos2013/impl.tex

Events are created on demand and are {\em owned} by the node on which they
were created.  The creation of an event occurs with no inter-node communcation.
The space of event
handles is divided across the nodes at startup time, allowing each node to assign handles
to new events without the risk of conflicting with another node's assignments.  The static division
of event handles permits any node to determine the owning node of an event
without any communication.
%No broadcast of
%the event creation is required either, as the static division is sufficient to allow any
%node to correctly determine an event's owning node when (and only if) an operation on that
%event is performed on the other node.

%At event creation time, the owning node allocates a data structure to track the state of the event
%(i.e. triggered or not) as well as record a list of the operations that are known to be dependent
%on the event (e.g. initiation of a copy operation, placing a task on a processor's ready queue, waking
%up a waiting task).  However, the owning node only keeps a list of the dependent operations from the same
%node.  Every other node also allocates a corresponding data structure (the first time the event
%is referenced) to remember if the event has already triggered and to record its local dependent operations.
%Arbitrarily many dependent operations from a
%single node are aggregated into a single {\em event subscription} active message that is sent to the
%owning node.  The owning node keeps a bitmask of which other nodes have subscribed to the event, and when the
%event finally does trigger, a single {\em event trigger} active message is sent to each subscribing node,
%which notes that the event has triggered (in case queries come after the trigger) and executes the list
%of local operations that were dependent on that event.  (In the case that the triggering of the event has
%occurred before the reception of an event subscription, a trigger message is sent immediatel to the 
%new subscriber.)
When a new event is created, the owning node allocates a data structure to track the state of
the event ({\em triggered} or {\em untriggered}) as well as to record a list of operations dependent on the
event from the same node called {\em waiters} (e.g. copy operation or task launch).  The first operation contingent on an event
from a remote node will create the same data structure on the remote node.  A single
{\em event subscription} active message is then sent from the remote node to the owning node indicating
that the remote node should be informed when the event triggers.  Arbitrarily many additional operations on
the remote node can then be aggregated as waiters onto the remote data structure without any additional communication.
When the event does trigger, the owner node notifies all local waiters and
sends a single {\em event trigger} active message to all nodes from which it has seen an event subscription
to notify remote waiters.
In the case where an event triggers while a subscription message is in flight from a remote node, the owner node will
immediately respond with a trigger message.

The actual triggering of an event may occur on any node.  If it occurs on a node other than the owning
node, an {\em event trigger} active message is sent from the triggering node back to the owning node, which
then forwards that message to all the other subscribed nodes.  The triggering node will automatically
notify its waiters and no message is sent from the owning node back to the triggering node.
%with the exception of the triggering node, which was
%able to execute its locally dependent operations (if any) immediately.  
While a remote trigger of an event can result in
the latency of a triggering operation being two active messages, it bounds the number of active
messages required per event to $2N-2$ where $N$ is the number of nodes 
%(which can be much smaller than the number of dependent operations) 
interested in the event.  An alternative
would have been to share the list of subscribers so that the triggering node could notify all interested
nodes directly.  However, such an algorithm is both more complicated due to race conditions, and requires $O(N^2)$
active messages.  As we will see in Section~\ref{subsec:eventmicro}, the latency of a single event trigger
active message is very small, and any algorithm that is super-linear with the node count will not scale well.
%either is or will soon be an scalability issue for large systems.

@@

%Instead of focusing on freeing event data structures, our implementation aggressively recycles them, 
%needing fewer total event data structures than a referencing counting implementation while also eliminating
%the overhead of reference counting.  The key observation is that one {\em generational event} data structure
% can efficiently capture the state of one {\em untriggered} event and a very large number (e.g. $2^{32}-1$)
%of already-triggered events.  An event's handle is expanded to include its generation number as well as the
%identifier for the underlying generational event.  In addition to the triggered-or-not state and list of
%dependent operations
%for the most recent generation, the owning node also remembers how many previous generations have already
%been triggered.  (There is no need to keep a list of dependent operations on previous generations.  Any
%new operation that is dependent on a generation that is known to be already triggered can immediately be
%executed.)  When an event is created, any generational event for whom the most recent generation has 
%triggered can be reused - the generation is increased by one and the state is returned to ``untriggered.''
%As before, this can be done with no inter-node communication.
Instead of attempting to free event data structures, our implementation aggressively recyles them.
Compared to a reference counting scheme our implementation requires fewer total event data structures
and has no programmer/runtime overhead.  The key observations is that one {\em generational event}
data structure can efficiently capture the state of one untriggered event and a very large 
number (e.g. $2^{32}-1$) of already-triggered events.  To accomplish this, we extend each event
handle to include its generation number as well as the identifier for its generational event.  Each
generational event remembers how many previous generations have already been triggered.  Any
new operation that is dependent on a generation that is known to already be triggered can immediately
be executed.  A generational event can be recycled as soon as the current generation has triggered. To
create a new event, a node finds a generational event in the triggered state,
increases the generation by one, and sets the generational event's state to untriggered.  As before, 
this can be done with no communication.

%A remote node's data structure is also efficient - the boolean values for whether the event
%has triggered and whether the event has already been subscribed to are replaced with the numbers of the most
%recent generation known to have triggered (this is received in the event trigger active message) and of the
%most recent generation the remote node has subscribed to (this is sent in the event subscription active
%message).  And as with the owner node's data structure, only
%one list of dependent operations is needed.  Although the latencies of a large system can delay the
%reception of an event trigger active message, if an operation is created that depends on a later generation
%of an event than what the current list of events is dependent on, that serves as a roundabout indication that
%the previous generation has indeed triggered, allowing the existing list of operations to be executed.
Nodes also maintain generational event data structures corresponding to remote events that they have
observed.  These data structures maintain the most recent known generation to have triggered as well
as the generation of the most recent subscription message sent (if any).  The distributed nature
of the system allows remote generational events to perform an interesting optimization.  If at any point a
remote generational event observes a request to wait on a later generation than its current generation,
then it is safe to infer that all generations up to the requested generation have already 
been triggered because a new generation of the event was already created by the event's owner node.
All local waiters can then be notified even before receiving an event triggered message for the current generation.

Barriers are implemented as an extension of events.  In place of an event's triggered/untriggered state,
a barrier's owner tracks the number of outstanding arrivals.  When the arrival count 
reaches zero, the barrier is considered to have triggered.  Remote nodes send {\em barrier update}
active messages containing an adjustment value that is positive for {\tt alter\_arrival\_count} operations 
and negative for {\tt arrive} operations.  A version of vector clocks\cite{Fidge1998} is used
to detect the race condition in which the barrier owner receives an {\tt arrive} operation before the
corresponding {\tt alter\_arrival\_count}.  When detected, the decrement of the arrival count 
is delayed to avoid a spurious triggering of the barrier.

@@

The unfairness that results from a lock favoring local waiters over remote waiters is intentional.  When the
contention on a lock is high (the only time when the question of fairness is relevant), the latencies
involved in transferring a lock between nodes can be the limiter on throughput.  By minimizing the number
of lock transfers, the throughput on the lock is maximized.

@@

In addition to normal physical regions, our implementation supports two different layouts of data for
reduction-only physical regions called {\em reduction instances}.   
These layouts are designed to optimize for the common pattern where 
multiple tasks make commutative {\em reductions} into shared data.  Instead of having to copy the existing 
physical region to a suitable memory and then applying individual reduction operations, reduction instances 
allow a large number of reduction operations to be accumulated and transferred efficiently to the original 
location of the data. This decoupling from the original location of the data can allow allow greater parallelism
in cases where processors cannot all access the same memory (i.e. many GPUs).

>>>>

Events are created on demand and are {\em owned} by the node on which they
were created.  The creation of an event occurs with no inter-node communcation.
The space of event
handles is divided across the nodes at startup time, allowing each node to assign handles
to new events without the risk of conflicting with another node's assignments.  The static division
of event handles also permits any node to determine the owning node of an event
without any communication.
%No broadcast of
%the event creation is required either, as the static division is sufficient to allow any
%node to correctly determine an event's owning node when (and only if) an operation on that
%event is performed on the other node.

%At event creation time, the owning node allocates a data structure to track the state of the event
%(i.e. triggered or not) as well as record a list of the operations that are known to be dependent
%on the event (e.g. initiation of a copy operation, placing a task on a processor's ready queue, waking
%up a waiting task).  However, the owning node only keeps a list of the dependent operations from the same
%node.  Every other node also allocates a corresponding data structure (the first time the event
%is referenced) to remember if the event has already triggered and to record its local dependent operations.
%Arbitrarily many dependent operations from a
%single node are aggregated into a single {\em event subscription} active message that is sent to the
%owning node.  The owning node keeps a bitmask of which other nodes have subscribed to the event, and when the
%event finally does trigger, a single {\em event trigger} active message is sent to each subscribing node,
%which notes that the event has triggered (in case queries come after the trigger) and executes the list
%of local operations that were dependent on that event.  (In the case that the triggering of the event has
%occurred before the reception of an event subscription, a trigger message is sent immediatel to the 
%new subscriber.)
When a new event is created, the owning node allocates a data structure to track the state of
the event ({\em triggered} or {\em untriggered}) as well as to record a list of operations dependent on the
event from the same node called {\em waiters} (e.g. copy operation or task launch).  The first operation contingent on an event
from a remote node will create the same data structure on the remote node.  A single
{\em event subscription} active message is then sent from the remote node to the owning node indicating
that the remote node should be informed when the event triggers.  Arbitrarily many additional operations on
the remote node can then be aggregated as waiters onto the remote data structure without any additional communication.
When the event does trigger, the owner node notifies all local waiters and
sends a single {\em event trigger} active message to each node from which it has seen an event subscription
to notify remote waiters.
In the case where an event triggers while a subscription message is in flight from a remote node, the owner node will
immediately respond with a trigger message.

The actual triggering of an event may occur on any node.  If it occurs on a node other than the owning
node, an {\em event trigger} active message is sent from the triggering node back to the owning node, which
then forwards that message to all the other subscribed nodes.  The triggering node will automatically
notify its waiters and no message is sent from the owning node back to the triggering node.
%with the exception of the triggering node, which was
%able to execute its locally dependent operations (if any) immediately.  
While a remote trigger of an event can result in
the latency of a triggering operation including two active message flight times, it bounds the number of active
messages required per event to $2N-2$ where $N$ is the number of nodes 
%(which can be much smaller than the number of dependent operations) 
interested in the event.  An alternative
would have been to share the list of subscribers so that the triggering node could notify all interested
nodes directly.  However, such an algorithm is both more complicated due to race conditions, and requires $O(N^2)$
active messages.  As we will see in Section~\ref{subsec:eventmicro}, the latency of a single event trigger
active message is very small, and any algorithm that is super-linear with the node count will not scale well.
%either is or will soon be an scalability issue for large systems.

@@

%Instead of focusing on freeing event data structures, our implementation aggressively recycles them, 
%needing fewer total event data structures than a referencing counting implementation while also eliminating
%the overhead of reference counting.  The key observation is that one {\em generational event} data structure
% can efficiently capture the state of one {\em untriggered} event and a very large number (e.g. $2^{32}-1$)
%of already-triggered events.  An event's handle is expanded to include its generation number as well as the
%identifier for the underlying generational event.  In addition to the triggered-or-not state and list of
%dependent operations
%for the most recent generation, the owning node also remembers how many previous generations have already
%been triggered.  (There is no need to keep a list of dependent operations on previous generations.  Any
%new operation that is dependent on a generation that is known to be already triggered can immediately be
%executed.)  When an event is created, any generational event for whom the most recent generation has 
%triggered can be reused - the generation is increased by one and the state is returned to ``untriggered.''
%As before, this can be done with no inter-node communication.
Instead of attempting to free event data structures, our implementation aggressively recycles them.
Compared to a reference counting scheme our implementation requires fewer total event data structures
and has no programmer/runtime overhead.  The key observations is that one {\em generational event}
data structure can efficiently capture the state of one untriggered event and a very large 
number (e.g. $2^{32}-1$) of already-triggered events.  To accomplish this, we extend each event
handle to include its generation number as well as the identifier for its generational event.  Each
generational event remembers how many previous generations have already been triggered.  Any
new operation that is dependent on a generation that is known to already be triggered can immediately
be executed.  A generational event can be reused for a new generation as soon as the current generation has triggered. To
create a new event, a node finds a generational event in the triggered state,
increases the generation by one, and sets the generational event's state to untriggered.  As before, 
this can be done with no inter-node communication.

%A remote node's data structure is also efficient - the boolean values for whether the event
%has triggered and whether the event has already been subscribed to are replaced with the numbers of the most
%recent generation known to have triggered (this is received in the event trigger active message) and of the
%most recent generation the remote node has subscribed to (this is sent in the event subscription active
%message).  And as with the owner node's data structure, only
%one list of dependent operations is needed.  Although the latencies of a large system can delay the
%reception of an event trigger active message, if an operation is created that depends on a later generation
%of an event than what the current list of events is dependent on, that serves as a roundabout indication that
%the previous generation has indeed triggered, allowing the existing list of operations to be executed.
Nodes also maintain generational event data structures corresponding to remote events that they have
observed.  These data structures maintain the most recent known generation to have triggered as well
as the generation of the most recent subscription message sent (if any).  The distributed nature
of the system allows remote generational events to perform an interesting optimization.  If at any point a
remote generational event receives a request to wait on a later generation than its current
generation,
then it is safe to infer that all generations up to the requested generation have already 
been triggered because a new generation of the event was already created by the event's owner node.
All local waiters can then be notified even before receiving the event trigger message for the current generation.

Barriers are implemented as an extension of events.  In place of an event's triggered/untriggered state,
a barrier's owner tracks the number of outstanding arrivals.  When the arrival count 
reaches zero, the barrier is considered to have triggered.  Remote nodes send {\em barrier update}
active messages containing an adjustment value that is positive for {\tt alter\_arrival\_count} operations 
and negative for {\tt arrive} operations.  A simplified version of vector clocks\cite{Fidge1998} is used
to detect the race condition in which the barrier owner receives an {\tt arrive} operation before the
corresponding {\tt alter\_arrival\_count}.  When detected, the decrement of the arrival count 
is delayed to avoid a spurious triggering of the barrier.

@@

The unfairness that results from a lock favoring local waiters over remote waiters is intentional.  When the
contention on a lock is high (the only time when the question of fairness is relevant), the latencies
involved in transferring a lock between nodes can be the limiter on throughput.  By minimizing the number
of lock transfers, the throughput on the lock is maximized.  This effect will be demonstrated in
Section~\ref{subsec:lockmicro}.

@@

In addition to normal physical regions, our implementation supports two different layouts of data for
reduction-only physical regions called {\em reduction instances}.   
These layouts are designed to optimize for the common pattern where 
multiple tasks make commutative {\em reductions} into shared data.  Instead of having to copy the existing 
physical region to a suitable memory and then applying individual reduction operations, reduction instances 
allow a large number of reduction operations to be accumulated and transferred efficiently to the original 
location of the data. This decoupling from the original location of the data can allow allow greater parallelism
in cases where processors cannot all access the same memory (e.g. many GPUs).
